---
pagetitle: "1- Data Preprocessing"
title: "Week 1: Data Preprocessing"
subtitle: Big Data and Machine Learning for Applied Economics | [MECA-4107](https://bloqueneon.uniandes.edu.co/d2l/home/130328)
author: 
      name: Eduard-Martínez
      affiliation: Universidad de los Andes  #[`r fontawesome::fa('globe')`]()
# date: Lecture 10  #"`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document:
    theme: flatly
    highlight: haddock
    # code_folding: show
    toc: yes
    toc_depth: 4
    toc_float: yes
    keep_md: false
    keep_tex: false ## Change to true if want keep intermediate .tex file
    ## For multi-col environments
  pdf_document:
    latex_engine: xelatex
    toc: true
    dev: cairo_pdf
    # fig_width: 7 ## Optional: Set default PDF figure width
    # fig_height: 6 ## Optional: Set default PDF figure height
    includes:
      in_header: tex/preamble.tex ## For multi-col environments
    pandoc_args:
        --template=tex/mytemplate.tex ## For affiliation field. See: https://bit.ly/2T191uZ
always_allow_html: true
urlcolor: blue
mainfont: cochineal
sansfont: Fira Sans
monofont: Fira Code ## Although, see: https://tex.stackexchange.com/q/294362
## Automatically knit to both formats
---

## **0. Initial configuration**

Require the libraries of this class:

```{r}
## install pacman
if(!require(pacman)) install.packages("pacman") ; require(pacman)

## require/install packages on this session
require(pacman)
p_load(rio, # import/export data
       tidyverse, # tidy-data
       skimr, # summary data
       caret) # Classification And REgression Training

## set seed
set.seed(0000)
```

## **1. load data**

First we are going to import some part of the GEIH data that will be used in the **problem-set 1**. The data dictionary is available [here](https://ignaciomsarmiento.github.io/GEIH2018_sample/dictionary.html)

```{r}
## load data
db <- import("https://gitlab.com/Lectures-R/bd-meca-2022-summer/lecture-01/-/raw/main/data/GEIH_sample1.Rds")
```

#### **1.1. dataframe vs tibble**

**Wha is a tibble?:** Tibbles are a modern take on data frames. They keep the features that have stood the test of time, and drop the features that used to be convenient but are now frustrating (i.e. converting character vectors to factors).

Then, **Why should I use them?:** There are three key differences between tibbles and data frames: printing, subsetting, and recycling rules. Run `vignette("tibble")` on the console to read more about it.

```{r}
db <- as_tibble(db) ## from dataframe to tibble
```

#### **1.2. Inspect data**

You can print in console both first or last database observations:

```{r}
## print data
head(db)
tail(db)
```

You can get a complete description of the database using the `skim()` function of the `skimr` library. However, you can to produce result summaries of the a variable using `summary()` function. Let me show you a example:

```{r}
## summary db
skim(db) %>% head()

## summary var
summary(db$y_salary_m)
```

## **2. visualize data**

```{r}
## data + mapping
ggplot(data = db , mapping = aes(x = age , y = y_ingLab_m))
```


```{r}
## + geometry
ggplot(data = db , mapping = aes(x = age , y = y_ingLab_m)) +
geom_point(col = "red" , size = 0.5)
```


```{r}
## by group
ggplot(data = db , 
       mapping = aes(x = age , y = y_ingLab_m , group=as.factor(formal) , color=as.factor(formal))) +
       geom_point()
```

You can save the chart in a object:

```{r}
## density: income by sex
p <- ggplot(data=db) + 
     geom_histogram(mapping = aes(x=y_ingLab_m , group=as.factor(sex) , fill=as.factor(sex)))
p
```

And you can add attributes to the object that contains the chart:

```{r}
p + scale_fill_manual(values = c("0"="red" , "1"="blue") , label = c("0"="Hombre" , "1"="Mujer") , name = "Sexo")
```

Boxplot chart:

```{r}
## box_plot: estrato1 vs totalHoursWorked
box_plot <- ggplot(data=db , mapping = aes(as.factor(estrato1) , totalHoursWorked)) + 
            geom_boxplot() 
box_plot
```

```{r}
## add another geometry
box_plot <- box_plot +
            geom_point(aes(colour=as.factor(sex))) +
            scale_color_manual(values = c("0"="red" , "1"="blue") , label = c("0"="Hombre" , "1"="Mujer") , name = "Sexo")
box_plot
```

```{r}
## add theme
box_plot + theme_test()
```

## **3. transformations**

#### **3.1 centering and scaling**

```{r}
h_hour = ggplot() + geom_histogram(data=db , aes(x=hoursWorkUsual) , fill="#99FF33" , alpha=0.5)
h_hour
```

```{r}
db = db %>% mutate(esc_hoursWorkUsual = scale(hoursWorkUsual))

h_hour + geom_histogram(data=db , aes(x=esc_hoursWorkUsual) , fill="#FF0066" , alpha=0.5)
```

#### **3.2 Skewness: log**

```{r,eval=F}
BoxCoxTrans(db$y_ingLab_m)
```

**Aaaa! We'll deal with NAs in a while, let's avoid them for now!**

```{r}
BoxCoxTrans(db$y_ingLab_m , na.rm=T)
```

```{r}
ggplot() + geom_boxplot(data=db ,aes(x=y_ingLab_m) , fill="darkblue" , alpha=0.5)
```

```{r}
db <- db %>% mutate(log_ingLab_m=log(y_ingLab_m))

ggplot() + geom_histogram(data=db , aes(x=log_ingLab_m) , fill="coral1" , alpha=0.5)
```

#### **3.3 resolve outliers **

```{r}
quantile(x=db$p6426 , na.rm=T)

IQR(x=db$p6426 , na.rm=T)

iqr <- IQR(x=db$p6426 , na.rm=T)

db_out <- db %>% subset(p6426 <= 2*iqr | is.na(p6426)==T)

cat("¡Elimina las NA!")

quantile(x=db_out$p6426 , na.rm=T)

nrow(db) - nrow(db_out)
```

## **4. Create, modify, and delete columns**


#### **4.1 dummy vars **

```{r}
db = db %>% 
     mutate(p6426_out = ifelse(test = p6426 > 4*iqr, 
                               yes = 1, 
                               no = 0))
table(db$p6426_out)
```

#### **4.2 multiple categorical var**

```{r}
q = quantile(db$p6426 , na.rm=T)
q
db = db %>% 
     mutate(p6426_q = case_when(p6426 < q[2] ~ "Q-1", 
                                p6426 >= q[2] & p6426 < q[3] ~ "Q-2", 
                                p6426 >= q[3] & p6426 < q[4] ~ "Q-3", 
                                p6426 >= q[4] ~ "Q-4"))
table(db$p6426_q)
```

#### **4.3 keeping only some vars**
```{r}
## select: delete a variable
head(db)

db %>% select(-Var.1,-orden)
```

```{r}
## select variable: by patter name 
db %>% select(starts_with("p6"))

db %>% select(directorio,contains("salary"))
```

```{r}
## select variable: by class
db %>% select_if(is.character)
```

## **5. dealing with missing values**

```{r}
## get missing values
is.na(db$y_total_m) %>% table()
```

```{r}
## replace values:
db %>% select(directorio,y_total_m) %>% tail()

db = db %>% 
     group_by(directorio) %>% 
     mutate(mean_y_total_m = mean(y_total_m,na.rm=T))

db %>% select(directorio,y_total_m,mean_y_total_m) %>% tail()
```

```{r}
db = db %>%
     mutate(y_total_m = ifelse(test = is.na(y_total_m)==T,
                               yes = mean_y_total_m,
                               no = y_total_m))

db %>% select(directorio,y_total_m,mean_y_total_m) %>% tail()
```


```{r}
## delete observations
db_c = db %>% subset(is.na(y_total_m)==F)

nrow(db) - nrow(db_c)

is.na(db$y_total_m) %>% table()

db_c = db %>% dplyr::filter(is.na(y_total_m)==F)
```

### **References**

-   Colin Gillespie and Robin Lovelace, 2017. Efficient R Programming, A
    Practical Guide to Smarter Programming [[Avaiable here]](https://csgillespie.github.io/efficientR/)

    *   Cap. 3: Efficient programming
    
    *   Cap. 4: Efficient programming
    
    
-   Kuhn, M., & Johnson, K. (2013). Applied predictive modeling (Vol.
    26, p. 13). New York: Springer. [[Avaiable here]](https://books.google.com.co/books/about/Applied_Predictive_Modeling.html?id=xYRDAAAAQBAJ&redir_esc=)

    *   Cap. 3: Data Pre-processing



